plyranges: a grammar for transforming genomics data

### 2018-03-26 paper.Rmd
restructure of design section

$$
\textit{GenomicRange} := \{  \{\text{seqname}_i, \left[\text{start}_i, \text{end}_{i}\right] , \text{strand}_i\}  : i = 1\dots n \}.
$$
A _GenomicRanges_ object can have an additional property that provides
annotation for each sequence name over a genome. This contstrains the the 
start and end points for a given sequence to lie inside the bounds
of the genome space. Additional values can be bound to the bare _IntegerRanges_ or _GenomicRanges_ and form metadata columns.

By definition the _Ranges_ class encapsulate tidy data prinicples because
every row corresponds to a sequence and every column corresponds to a variable.
Hence, we have designed the `plyranges` DSL to have the grammar
and design principles of `dplyr`: fluency, consistency, cohesion and endomorphism.

To promote fluency we have designed every function call to perform an
action on a _Ranges_ object and where possible
functions have few arguments. Each function is named after a verb and is designed
to enable a tab completion based workflow. Both of these aspects reduce the 
cognitive load on a new user since most manipulations can be performed with a 
vocabulary of several verbs, rather than having to memorise functions (that
are named as nouns) with many arguments (which is required in the existing Bioconductor packages). Workflows can be composed by chaining verbs together into
'sentences' via the forward pipe operator,`%>%` (exported from the R package `magrittr` @R-magrittr). Overall, this allows users to write human readable code
because workflows describe what the code is doing rather than how its doing it.

In order to provide a compatible API with _dplyr_, _plyranges_ makes extensive 
use of non-standard evaluation in R via the _rlang_ package @R-rlang. Simply, this
means that computations are performed and evaluated in the context of the 
_Ranges_ objects, which emphasises the interactive nature of the API. This has
the disadvantage that programming with _plyranges_ becomes more difficult 
because a user needs to capture expressions inside function calls and then
unquote them.

 This is possible because every 
function call is an endomorphism: when the input is _Ranges_ object the output
will also be a _Ranges_ object. One advantage of endomorphism is that it 
does not  require any additional learning of classes beyond _Ranges_ and 
the _DataFrame_ classes. This strongly deviates from the design of 
the _Ranges_ Bioconductor packages, where many methods return a new class
upon return. The Bioconductor design enables efficient computing as users
are exposed to low-level features of its API which _plyranges_ abstracts
away. Method chaining via the pipe operator can also be difficult to debug,
as there multiple points of failure.

<!--- need to make this clearer? --->

## Actions on Ranges

The _plyranges_ API exports the six core verbs from the _dplyr_ package
and modifies them for use with _Ranges_ objects. The
verb `mutate()` takes a Ranges object and a set of name-value pairs
and generates a new Ranges object that with modified or new metadata
columns or modified core components (start, end, width, seqnames, strand).
The use of `mutate()` means that a user no longer needs knowledge of the 
accessors of the Ranges object, as they can modify them in place.
The `filter()` function takes a Ranges object and logical expressions and
restricts Ranges object to where the logical expression evaluates to true.
The `summarise()` function takes a Ranges object and a set of name-value pairs
and aggregates the Ranges according to functions evaluated in the name-value pairs.
As `summarise()` is an aggregation it may break the structure of the of a Ranges
object, hence it returns a _DataFrame_ object.
The `select()` function determines
which metadata columns are returned and the order they are returned in. The 
`arrange()` function sorts a Ranges object by named variables. 
The `group_by()` function creates an implicit grouping of Ranges object according
to variables in the Ranges object. This modifies the actions of `mutate()`,
`summarise()` and `filter()`, so they are performed on each partition created 
by the grouping. The `group_by()` operation acts as a replacement for 
the _GenomicRangesList_ class in the original _GenomicRanges_ API.

The _plyranges_ API introduces additional summary verbs, `reduce_ranges()`
and `disjoin_ranges()`that return Ranges objects after being returned. The
`reduce_ranges()` operation merges overlapping and neighbour ranges, while
`disjoin_ranges()` expands ranges by taking the union of end points.

## Arithimetic on Ranges

@@

@@

The combination of verbs we have defined above encapsulate all operations that
can be performed in the original _IRanges_ and _GenomicRanges_ packages
without the user be exposed to new classes. In those packages to perform
most operations requires users to learn many classes and perform 
additional manipulations to return the results of their computation back to a
_Ranges_ object.

## Construction and Import/Output

>>>>

$$
\textit{GenomicRange} := \{  \{\text{seqname}_i, \left[\text{start}_i, \text{end}_{i}\right] , \text{strand}_i\}  : i = 1\dots n \}.
$$
A _GenomicRanges_ object can have an additional property that provides
annotation for each sequence name over a genome. This contstrains the the 
start and end points for a given sequence to lie inside the bounds
of the genome space. Additional values can be bound to the bare 
_IntegerRanges_ or _GenomicRanges_ to form metadata columns.

By definition the _Ranges_ class encapsulate tidy data prinicples because
every row corresponds to a sequence and every column corresponds to a variable.
Hence, we have designed the `plyranges` DSL to have the grammar
and design principles of `dplyr`: fluency, consistency, cohesion and 
endomorphism. All of these principles are discussed with respect to the 
_Ranges_ data model below. Where possible we contrast our design to the existing
Bioconductor infrastructure.

A caveat to constructing a compatible API with `dplyr` is that 
`plyranges` makes extensive use of non-standard evaluation in R
(achieved via the `rlang` package @R-rlang). Simply, this
means that computations are performed and evaluated in the context of the 
_Ranges_ objects, which emphasises the interactive nature of the API. This has
the disadvantage that programming with `plyranges` becomes more difficult 
because a user needs to know about non-standard evaluation and user defined
functions can be difficult to debug.

## Fluency

We have designed every function to perform a single
action on a _Ranges_ object. The `plyranges` DSL implments the core verbs from the _dplyr_ package
and implements context specific verbs for transforming _Ranges_ objects (described in
table ). Each verb preserves the semantics of _Ranges_ object and via
duck typing works with derivatives of the _Ranges_ class. 
Both of these aspects reduce the cognitive load on a new user since most 
manipulations can be performed with a  vocabulary of several verbs, rather 
than having to memorise function names that are nouns. Workflows can be composed
by chaining verbs together into 'sentences' via the forward pipe operator,`%>%` 
(exported from the R package `magrittr` @R-magrittr), which can be read as the 
word 'then'. Overall, this  allows users to write human readable code because 
workflows describe what the code is doing rather than how its doing it.

```{r, echo = FALSE, message = FALSE, warning = FALSE}
library(tibble)
feature_tbl <- tribble(
  ~Verb, ~Action, ~Description,
  "`mutate()`", "modifies columns", "Takes a _Ranges_ object and a set of name-value pairs and generates a new _Ranges_ object that with modified or new metadata columns or modified core components (start, end, width, seqnames, strand).",
 "`filter()`", "subset rows", "Takes a _Ranges_ object and a set of logical expressions and restricts the _Ranges_ object to where all logical expressions are true.",
 "`summarise()`", "aggregate one or more columns by a function", "Takes a _Ranges_ object and a set of name-value pairs and aggregates the _Ranges_ according to functions evaluated in the name-value pairs.",
"`group_by()`", "partition a _Ranges_ object by a column", "Create an implicit grouping of _Ranges_ object by one or more columns. This modifies the actions of verbs, so they are performed on each partition created by the grouping.",
"`select()`", "select columns of a _Ranges_ objects", "Drop or reorder metadata columns.",
"`arrange()`", "sort a _Ranges_ object by columns", "Sort a _Ranges_ object by named variables",
"`disjoin_ranges()`", "aggregate one more columns of a _Ranges_ object over the union of end coordinates", "Takes a _Ranges_ object and a set of name-value pairs and expands the _Ranges_ object by taking the union of end coordinates. For each row that was included to form the final union, the aggregation function defined by a name-value pair is called.",
"`reduce_ranges()`", "aggregate one more columns of a _Ranges_ object by merging overlapping and neighbouring ranges.", "Takes a _Ranges_ object and a set of name-value pairs and reduces the _Ranges_ object by merging overlapping and neighouring ranges. For each row that was included to form merged range, the aggregation function defined by a name-value pair is called."
)
```

## Consistency

@@

## Endomorphism

 (that
are named as nouns) with many arguments (which is required in the existing 
Bioconductor packages)
 This is possible because every 
function call is an endomorphism: when the input is _Ranges_ object the output
will also be a _Ranges_ object. One advantage of endomorphism is that it 
does not  require any additional learning of classes beyond _Ranges_ and 
the _DataFrame_ classes. This strongly deviates from the design of 
the _Ranges_ Bioconductor packages, where many methods return a new class
upon return. The Bioconductor design enables efficient computing as users
are exposed to low-level features of its API which _plyranges_ abstracts
away. Method chaining via the pipe operator can also be difficult to debug,
as there multiple points of failure.

The combination of verbs we have defined above cover most operations that
can be performed in the original _IRanges_ and _GenomicRanges_ packages
without the user being exposed to new classes. In those packages to perform
most operations requires users to learn many classes and perform 
additional manipulations to return the results of their computation back to a
_Ranges_ object.

## Cohesion

@@
